Insurance choice, decision assistance, and insurer steering

### 2021-06-18 finals/ashecon/ashecon-202106.Rmd
Update slides for ashecon

---
title: "Insurance choice, decision assistance, and insurer steering"
subtitle: "Ian McCarthy & Evan Saltzman"
author: Carolina Region Empirical Economics Day
date: May 3, 2021 #"`r format(Sys.time(), '%d %B %Y')`"
header-includes: 
  - \usepackage{graphicx}
  - \usepackage{amsmath}
output:
#  html_document: default (toggle on for "simplified" view)
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts, custom.css] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      beforeInit: ["macros.js","https://platform.twitter.com/widgets.js"]
#knit: pagedown::chrome_print
---

@@

if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, ggplot2, dplyr, lubridate, readr, readxl, hrbrthemes,
               scales, gganimate, gapminder, gifski, png, tufte, plotly, OECD,
               ggrepel, xaringanExtra)
```

@@

# Some decisions are hard

.center[
<iframe width="800" height="500" src="https://www.youtube.com/embed/oWXNf3JVWpI?start=146&end=272" frameborder="0" allowfullscreen></iframe>
]

@@

1. Does some form of decision assistance "help" in plan selection?

2. Do insurers use brokers/agents to steer people to different plans?

3. What are the welfare effects of any such steering?

@@

1. [Data](#data)
2. [Dominated choices](#dominated)
3. [Effects of decision assistance](#decision-assist)
4. [Insurer steering](#steering)

@@

---
# Insurers

@@

---
# Metal tier

@@

@@

- Linear probability model of dominated choices or switching
- Dummy variable for receiving any assistance
- Selection on observables assumption (for now)

---
# Selection on observables

.pull-left[
  ![](../../figures/cov_balance.png)
]

.pull-right[
  ![](../../figures/ps_assist_full.png)

@@

.center[
  ![:scale 50%](../../figures/dominated_choice_regression.png)
]

---
# Decision assistance and switching

.center[
  ![:scale 50%](../../figures/switching_regression.png)
]

@@

1. Estimate multinomial logit model among those without any decision assistance, $\hat{\theta}$

2. Use $\hat{\theta}$ to form predicted counterfactual, $\widehat{Pr[Y_{ij}^{0} | \text{Assist}]}$

3. Use predicted value in ATT estimate: $$\underbrace{Pr[Y_{ij}^{1} | \text{Assist}]}_{Observed} - \underbrace{\widehat{Pr[Y_{ij}^{0} | \text{Assist}]}}_{Predicted}$$

4. Aggregate to level of insurer and metal

- Clustered bootstrap, by year and rating area

# Analysis of insurer steering

---
# This is preliminary!

.center[
  ![:scale 50%](https://media.giphy.com/media/3o6Yg9rDXHL8Zrdlza/giphy.gif)
]

@@

**Calculate commission elasticities of demand**

@@

|                | Agent Users   | All Enrollees  |
| -------------- |:-------------:| --------------:|
| Overall        | 11.18         | 5.39           |
| Large Insurers | 11.94         | 5.75           |
| Small Insurers | 6.33          | 3.08           |

>>>>

---
title: "Insurance choice, decision assistance, and insurer steering"
subtitle: "Ian McCarthy & Evan Saltzman"
author: ASHEcon Annual Conference
date: June 23, 2021 #"`r format(Sys.time(), '%d %B %Y')`"
header-includes: 
  - \usepackage{graphicx}
  - \usepackage{amsmath}
  - \usepackage{booktabs}
  - \usepackage{caption}
output:
#  html_document: default (toggle on for "simplified" view)
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts, custom.css] 
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      beforeInit: ["macros.js","https://platform.twitter.com/widgets.js"]
#knit: pagedown::chrome_print
---

@@

if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, ggplot2, dplyr, lubridate, readr, readxl, hrbrthemes,
               scales, gganimate, gapminder, gifski, png, tufte, plotly, OECD,
               ggrepel, xaringanExtra, kableExtra, knitr, htmltools)
```

@@

# Decisions are hard

.center[
  ![](https://media.giphy.com/media/Lp9wng87dsScq2owim/giphy.gif)
]

@@

1. Does some form of decision assistance affect plan selection?

2. Does decision assistance "improve" plan selection?

3. Do insurers use brokers/agents to steer people to different plans?

4. (not today) What are the welfare effects?

@@

1. [Data](#data)
2. [Effects of decision assistance](#decision-assist)
3. [Dominated choices](#dominated)
4. [Insurer steering](#steering)

@@

---
# Households

```{r, include=FALSE}
load('final_summary.Rdata')
```

```{r sum-stats, echo=FALSE}
kable(final.sum.stats %>% filter(name %in% c("Household Size", "No. of Children",
                                             "Percent Black", "Percent White",
                                             "Percent Hispanic","below 138%", "between 138 and 250%",
                                             "between 250 and 400%", "above 400%",
                                             "Total HHs (millions)", "Total Uninsured (millions)")), 
      format="html",
      col.names = c("Variable","Assisted","Unassisted","Overall"),
      digits=c(0,2,2,2),
      booktabs=T) %>%
  pack_rows("Household Demographics", 1, 5) %>%  
  pack_rows("Income relative to FPL", 6, 9) %>%
  kable_styling(full_width=F,
                font_size=17)
```

---
# Plan choice: Insurers

@@

---
# Plan choice: Metal tiers

@@

1. Calculate predicted probability of decision assistance from logit model

.pull-left[
  ![:scale 80%](../../figures/ps_assist_full.png)
]

.pull-right[
  ![:scale 80%](../../figures/cov_balance.png)
]

1. Calculate predicted probability of decision assistance from logit model
2. Estimate weighted multinomial logit model among those without any decision assistance, $\hat{\theta}$, with weights $\frac{\hat{p}}{1-\hat{p}}$
3. Use $\hat{\theta}$ to form predicted counterfactual, $\widehat{Pr[Y_{ij}^{0} | \text{Assist}]}$
4. Use predicted value in ATT estimate: $$\underbrace{Pr[Y_{ij}^{1} | \text{Assist}]}_{Observed} - \underbrace{\widehat{Pr[Y_{ij}^{0} | \text{Assist}]}}_{Predicted}$$
5. Aggregate to level of insurer and metal

@@

1. Calculate predicted probability, $\hat{p}$, of decision assistance from logit model

2. Estimate weighted logit model of dominated choice among those without any decision assistance, $\hat{\beta}$, with weights given by  $\frac{\hat{p}}{1-\hat{p}}$

3. Use $\hat{\beta}$ to form predicted counterfactual, $\widehat{Pr[Y_{i}^{0} | \text{Assist}]}$

4. Use predicted value in ATT estimate: $$\underbrace{Pr[Y_{i}^{1} | \text{Assist}]}_{Observed} - \underbrace{\widehat{Pr[Y_{i}^{0} | \text{Assist}]}}_{Predicted}$$

@@

.center[
  ![:scale 50%](../../figures/dom_choice.png)
]

# *Preliminary* analysis of insurer steering

@@

@@

@@

|                | Agent Users   | All Enrollees  |
| -------------- |:-------------:| --------------:|
| Overall        | 7.70          | 3.69           |
| Large Insurers | 8.07          | 3.86           |
| Small Insurers | 5.30          | 2.57           |
